---
title: "P5 code"
output: html_document
---

# My focus for this assignment

Data dictionary: https://nhts.ornl.gov/tables09/CodebookBrowser.aspx

Filter for a particular trip purpose and/or by a particular population $\rightarrow$ predict the likelihood that a short trip (< 1.5mi) will take place by walking (mode choice)

* **Research question:** are adults (18-65) living in urban areas more likely to walk for short trips (< 1.5mi) to run errands compared to adults (18-65) living in rural areas?
* Filter the population to only include people ages 18-65
* Filter the WHYTO or WHYFROM variables to include the trip purposes of interest
  - 11: buy goods (groceries, clothes, appliances, gas)
  - 12: buy services (dry cleaners, banking, service a car, pet care)
  - 13: buy meals (go out for a meal, snack, takeout)
  - 14: other general errands (post office, library)
* Predictors
  - Trip distance (TRPMILES)
  - Gender (R_SEX)
  - Age (R_AGE) 
  - Urban/rural (URBRUR)
  - Population density of the trip destination (DBPPOPDN) 

# Setup

### Load libraries

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(here)
library(knitr)
library(srvyr)
library(tidycensus)
library(jtools)
```

### Load data

```{r, message = FALSE, warning = FALSE}
trips <- here("P5_mode_choice", "trippub.csv") |>
  read_csv(show_col_types = FALSE)
```

```{r, message = FALSE, warning = FALSE}
people <- here("P5_mode_choice", "perpub.csv") |>
  read_csv(show_col_types = FALSE)
```

# Data cleaning

### Filter the trip purposes

```{r, message = FALSE, warning = FALSE}
# Filter the dataset to only include the trip purposes I'm interested in studying (11, 12, 13, 14)
trip_purposes <- c(11, 12, 13, 14)

visit_trips <- trips |>
  filter(WHYTO %in% trip_purposes |
           WHYFROM %in% trip_purposes)
```

### Filter the population

```{r, message = FALSE, warning = FALSE}
# Filter the dataset to only include adults age 18-65
sr_visit_trips <- visit_trips |>
  filter(R_AGE >= 18 & R_AGE <= 65)
```

### Filter for short distance trips

```{r, message = FALSE, warning = FALSE}
short_sr_visit_trips <- sr_visit_trips |>
  filter(TRPMILAD < 1.5)

nrow(short_sr_visit_trips)
```

We have a sample of 79,511 trips after all the filtering

### Generate the outcome variable

```{r, message = FALSE, warning = FALSE}
# Create a binary outcome variable of whether or not the mode of transport was walking 

short_sr_visit_trips <- short_sr_visit_trips |>
  mutate(walk = TRPTRANS == "01")

short_sr_visit_trips |>
  mutate(Mode = factor(ifelse(walk, "Walk", "Other mode"),
                       levels = c("Walk", "Other mode"))) |>
  group_by(Mode) |>
  summarise(`Number of trips` = n()) |>
  mutate(`Percent of trips` = 
           paste0(round(100*`Number of trips`/sum(`Number of trips`)), "%")) |>
  kable()
```

#### Incorporate survey weights

We're incorporating survey weights to re-estimate the percent of short trips to run errands for adults 18-65 that take place via walking vs. another mode

```{r, message = FALSE, warning = FALSE}
short_sr_visit_trips |>
  srvyr::as_survey_design(weights = WTTRDFIN) |>
  mutate(Mode = factor(ifelse(walk, "Walk", "Other mode"),
                       levels = c("Walk", "Other mode"))) |>
  group_by(Mode) |>
  srvyr::survey_tally(vartype = "ci") |>
  mutate(`Estimated percent of trips` = 
           paste0(round(100*n/sum(n)),"%"),
         `Lower estimate (95% confidence)` = 
           paste0(round(100*n_low/sum(n)),"%"),
         `Upper estimate (95% confidence)` = 
           paste0(round(100*n_upp/sum(n)),"%")) |>
  select(Mode,
         `Estimated percent of trips`,
         `Lower estimate (95% confidence)`,
         `Upper estimate (95% confidence)`) |>
  kable()
```

# Examining and cleaning predictors

### Trip distance (TRPMILES)

```{r, message = FALSE, warning = FALSE}
# Remove NAs (coded as negative values)
sample_trips <- short_sr_visit_trips |>
  filter(TRPMILES >=0)

ggplot(sample_trips) +
  geom_histogram(aes(x = TRPMILES),
                 color = "gray",
                 fill = "cornsilk",
                 binwidth = 0.1) +
  scale_x_continuous(name = "Trip distance (miles)",
                     breaks = seq(0, 1.5, by=0.1)) +
  scale_y_continuous(name = "Number of trips in sample") +
  theme_minimal()
```

### Gender (R_SEX)

```{r, message = FALSE, warning = FALSE}
# Remove NAs (coded as negative values)
# Male = 0 FALSE; Female = 1 TRUE
sample_trips <- sample_trips |>
  filter(R_SEX > 0) |>
  mutate(R_SEX_logical = R_SEX == "02") 

table(sample_trips$R_SEX_logical) 
prop.table(table(sample_trips$R_SEX_logical))
```

### Age (R_AGE) 

```{r, message = FALSE, warning = FALSE}
ggplot(sample_trips) +
  geom_histogram(aes(x = R_AGE),
                 color = "gray",
                 fill = "cornsilk",
                 binwidth = 1) +
  scale_x_continuous(name = "Traveler's age (years)",
                     breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(name = "Number of trips in sample") +
  theme_minimal()
```

### Urban/rural (URBRUR)

```{r, message = FALSE, warning = FALSE}
# Remove NAs (coded as negative values)
# Urban = 0 FALSE; Rural = 1 TRUE
sample_trips <- sample_trips |>
  filter(URBRUR > 0) |>
  mutate(URBRUR_logical = URBRUR == "02") 

table(sample_trips$URBRUR_logical) 
prop.table(table(sample_trips$URBRUR_logical))
```

### Population density of the trip destination (DBPPOPDN) 

```{r, message = FALSE, warning = FALSE}
# Remove NAs (coded as negative values)
sample_trips <- sample_trips |>
  filter(DBPPOPDN > 0)

density_values <- c(50, 300, 750, 1500, 3000, 7000, 17000, 30000)

ggplot(sample_trips) +
  geom_histogram(aes(x = DBPPOPDN),
                 color = "gray",
                 fill = "cornsilk",
                 bins = 30) +
  scale_x_continuous(breaks = density_values,
                 labels = density_values,
                 transform = "log",
                 name = paste0("Population density of destination\n",
                               "block group (people per sq-mi)")) +
  scale_y_continuous(name = "Number of trips in sample") +
  theme_minimal()
```

### Final sample size after removing NAs

```{r}
nrow(sample_trips)
```

The final sample size is 79,398 trips after removing NAs

# Logistic regression model

We will estimate a logistic regression model to predict the likelihood that a trip in our cleaned/filtered dataset will take place by walking.

```{r, message = FALSE, warning = FALSE}
model <- glm(walk ~ 
               TRPMILES +
               R_SEX_logical +
               R_AGE +
               URBRUR_logical +
               DBPPOPDN,
             data = sample_trips,
             family = "binomial")

coeff_labels <- c("Trip distance (miles)" = "TRPMILES",
                  "Sex (female)" = "R_SEX_logicalTRUE",
                  "Age (years)" = "R_AGE",
                  "Rural" = "URBRUR_logicalTRUE",
                  "Block-group population density at destination" = "DBPPOPDN")

export_summs(model, 
             robust = "HC3", 
             coefs = coeff_labels,
             error_format = "(p = {p.value})",
             error_pos = "right",
             digits = 4)
```

# Visualizing the impact of each predictor

### Trip distance (TRPMILES)

Effect plot of the predicted probabilities of walking across the range of trip distances, holding all other predictors at their average (if continuous) or reference (if categorical) values

```{r, message = FALSE, warning = FALSE}
effect_plot(model, pred = "TRPMILES", interval = TRUE) +
  scale_x_continuous(name = "Trip distance (miles)",
                     breaks = seq(0, 1.5, by  =0.1)) +
  scale_y_continuous(name = "Probabilitity of walking",
                     breaks = breaks <- seq(0, 1, by = 0.1),
                     labels = paste0(breaks*100, "%")) + ggtitle("Effect of Trip Distance on Walking Probability")
```

### Gender (R_SEX_logical)

```{r, message = FALSE, warning = FALSE}
effect_plot(model = model, pred = "R_SEX_logical", interval = TRUE) +
  scale_y_continuous(name = "Probability of walking for a particular trip",
                     breaks = breaks <- seq(0, 1, by=0.01),
                     labels = paste0(breaks*100, "%")) +
  scale_x_discrete(name = paste0("Sex"),
                   labels = c("Male", "Female")) + ggtitle("Effect of Sex on Walking Probability")
```
  
### Age (R_AGE) 

Effect plot of the predicted probabilities of walking across the range of ages, holding all other predictors at their average (if continuous) or reference (if categorical) values

```{r, message = FALSE, warning = FALSE}
effect_plot(model, pred = "R_AGE", interval = TRUE) +
  scale_x_continuous(name = "Age (years)",
                     breaks = seq(0, 70, by  =10)) +
  scale_y_continuous(name = "Probabilitity of walking",
                     breaks = breaks <- seq(0, 1, by = 0.01),
                     labels = paste0(breaks*100, "%")) + ggtitle("Effect of Age on Walking Probability")
```

### Urban/rural (URBRUR)

```{r, message = FALSE, warning = FALSE}
effect_plot(model = model, pred = "URBRUR_logical", interval = TRUE) +
  scale_y_continuous(name = "Probability of walking for a particular trip",
                     breaks = breaks <- seq(0, 1, by=0.01),
                     labels = paste0(breaks*100, "%")) +
  scale_x_discrete(name = paste0("Household is located in what kind of area"),
                   labels = c("Urban", "Rural")) + ggtitle("Effect of Household Urban vs. Rural Location on Walking Probability")
```

### Population density of the trip destination (DBPPOPDN) 

```{r, message = FALSE, warning = FALSE}
effect_plot(model, pred = "DBPPOPDN", interval = TRUE) +
  scale_x_continuous(name = "Population density of trip destination",
                     breaks = seq(0, 30000, by  =5000)) +
  scale_y_continuous(name = "Probabilitity of walking",
                     breaks = breaks <- seq(0, 1, by = 0.1),
                     labels = paste0(breaks*100, "%")) + ggtitle("Effect of Destination Population Density on Walking Probability")
```

