---
title: "P3_code"
output: html_document
---

# 1. Setup

### References

Data dictionary: https://nhts.ornl.gov/tables09/CodebookBrowser.aspx

P3 assignment instructions: https://gsd-ses-5394-sp2025.github.io/examples/P3/P3

### Loading libraries

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
```

### Loading data

```{r}
# Loading household-level data from the 2017 National Household Travel Survey (NHTS)
hh_data <- read.csv("/Users/paigelee/Documents/SES 5394/ses5394/P2_vehicle_availability/data/hhpub.csv")

# Loading person-level data
person_data <- read.csv("/Users/paigelee/Documents/SES 5394/ses5394/P2_vehicle_availability/data/perpub.csv")

# Loading trip data
trip_data <- read.csv("/Users/paigelee/Documents/SES 5394/ses5394/P2_vehicle_availability/data/trippub.csv")
```

# 2. Feature selection and engineering

### Creating the response variable

Create a binary response variable with values of 1 if the trip destination purpose (WHYTO) was for medical trips, or 0 otherwise.

```{r}
trip_data$medical_trips_response <- ifelse(trip_data$WHYTO == 18, 1, 0)
table(trip_data$medical_trips_response)
```

There is a severe class imbalance in the response variable, so we'll need to perform weighted logistic regression later.

### Calculating weights for weighted logistic regression

Due to the big class imbalance in the response variable (`trip_data$medical_trips_response`), we need to perform weighted logistic regression since medical trips (what we're focusing on) are very rare in the dataset

```{r}
trip_data$weights <- ifelse(trip_data$medical_trips_response == 1, 
                            nrow(trip_data) / (2 * sum(trip_data$medical_trips_response == 1)),  # Weight for medical trips
                            nrow(trip_data) / (2 * sum(trip_data$medical_trips_response == 0)))  # Weight for non-medical trips
```

### Creating binary predictors for age groups

Create two binary predictors: 1. whether a person is ≥ 65 years old, 2. whether a person is < 18 years old. 

```{r}
person_data$is_65_older <- ifelse(person_data$R_AGE >= 65, 1, 0)
table(person_data$is_65_older)
```

```{r}
person_data$is_18_younger <- ifelse(person_data$R_AGE < 18, 1, 0)
table(person_data$is_18_younger)
```

### Merge trip data with person data 

```{r}
# Left join trip data with person data -> keep all records from trip data, find their matches with person data
# Merge on the unique (HOUSEID, PERSONID) pairs -> merge on two columns

trip_data_merged <- merge(x = trip_data, y = person_data, by = c("HOUSEID", "PERSONID"), all.x = TRUE)  
```

### Clean the merged data for duplicate columns .x and .y

```{r}
# Function to check if .x and .y versions are identical
check_identical_columns <- function(df) {
  cols_x <- grep(".x$", names(df), value = TRUE)  # Get .x columns
  cols_y <- sub(".x$", ".y", cols_x)  # Corresponding .y columns

  identical_cols <- c()  # To store columns where .x and .y are identical
  non_identical_cols <- c()  # To store columns where .x and .y are different

  for (i in seq_along(cols_x)) {
    col_x <- df[[cols_x[i]]]
    col_y <- df[[cols_y[i]]]

    # Check if they are identical (ignoring NAs)
    if (all(col_x == col_y, na.rm = TRUE)) {
      identical_cols <- c(identical_cols, cols_x[i])  # Store identical ones
    } else {
      non_identical_cols <- c(non_identical_cols, cols_x[i])  # Store non-identical ones
    }
  }

  list(identical = identical_cols, non_identical = non_identical_cols)
}

cat("Number of columns in trip_data_merged before cleaning:", ncol(trip_data_merged), "\n")

# Check which columns are identical
col_check <- check_identical_columns(trip_data_merged)

# Process only if columns are identical
trip_data_merged <- trip_data_merged %>%
  # Keep .x version if .x and .y are identical
  mutate(across(all_of(col_check$identical), ~ coalesce(.x, get(sub(".x$", ".y", cur_column()))))) %>%
  # Drop the corresponding .y columns
  select(-all_of(sub(".x$", ".y", col_check$identical))) %>%
  # Rename .x columns by removing .x suffix
  rename_with(~ sub(".x$", "", .), all_of(col_check$identical))

# Output non-identical columns for manual review
if (length(col_check$non_identical) > 0) {
  cat("Warning: These columns have different .x and .y values and were not merged:\n")
  print(sub(".x$", "", col_check$non_identical))
}

cat("Number of columns in trip_data_merged after cleaning:", ncol(trip_data_merged))
```

# 3. Modeling

### Model 1: weighted logistic regression with quadratic age (no age predictors) 

```{r, message = FALSE, warning = FALSE}
model1 <- glm(medical_trips_response ~ R_AGE + I(R_AGE^2), 
                      family = binomial(link = "logit"), 
                      data = trip_data_merged, 
                      weights = trip_data_merged$weights)

summary(model1)
```





