---
title: "P4_code"
output: html_document
---

# Load libraries

```{r, message = FALSE, warning = FALSE}
# Set Java 21 rather than Java 24
Sys.setenv(JAVA_HOME = "/Library/Java/JavaVirtualMachines/jdk-21.jdk/Contents/Home")
library(rJava)
options(java.parameters = '-Xmx4G')

library(tidyverse)
library(here)
library(knitr)
library(tigris)
library(stringr)
library(maptiles)
library(tidyterra)
library(r5r)
library(sf)
library(leaflet)

here("P4_trip_distribution", "grvty_balancing.R") |> source()
```

# Select a study area

There are 939 core-based statistical areas (CBSAs) in the U.S. $\rightarrow$ 393 are metropolitan areas (MSAs), and 542 are micropolitan areas ($\mu$SAs) 

```{r}
all_cbsas <- core_based_statistical_areas(progress_bar = FALSE,
                                          year = 2024) |>
  select(NAMELSAD) |>
  mutate(type = ifelse(!is.na(str_match(NAMELSAD, "Metro")), "Metro", "Micro")) |>
  mutate(type = as.character(type))

table(all_cbsas$type) |>
  kable()
```

The sweet spot is to select a large micro area or a small metro area.

```{r}
salem <- all_cbsas |>
  filter(NAMELSAD == "Salem, OH Micro Area") |>
  st_transform("WGS84")

base_map <- get_tiles(salem,
                      provider = "CartoDB.Positron",
                      zoom = 9,
                      crop = TRUE)

ggplot(salem) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(fill = NA,
          color = "orange") +
  theme_void()
```

# Load job data

The LEHD Origin-Destination Employment Statistics (LODES) dataset provides the number of workers who live and work between any pair of census blocks in a given state in a given year. 

Counties in the Salem, OH Micro Area

* Columbiana County, OH (FIPS code: 39029)

Since the Youngstown-Warren, OH Metro Area spans both OH and PA, we'll need to download and process the LODES data for both states, then combine the results and filter for the counties of interest.

```{r}
# Set year and FIPS codes
year <- "2021"
salem_counties_5_digit <- c("39029")  

# Function to load and filter data for a given state
load_state_data <- function(state_abbrev, counties_5_digit, year) {
  url <- paste0("https://lehd.ces.census.gov/data/lodes/LODES8/",
                state_abbrev,
                "/od/",
                state_abbrev,
                "_od_main_JT00_",
                year,
                ".csv.gz")
  
  read_csv(url, col_types = cols()) |>
    mutate(w_county = substr(w_geocode, 1, 5),
           h_county = substr(h_geocode, 1, 5)) |>
    filter(h_county %in% counties_5_digit & w_county %in% counties_5_digit) |>
    mutate(w_geocode = as.character(w_geocode),
           h_geocode = as.character(h_geocode))
}

# Load the OH LODES data 
oh_data <- load_state_data("oh", salem_counties_5_digit, year)

head(oh_data)
```

# Aggregate data to zone totals

There are three industry categories: goods, trade, and services. We want to create a trip generation table that shows the number of workers produced (by living in a zone) and attracted to (by working in a zone) for each industry category.

```{r}
# Calculate the total number of trip productions
total_prod <- oh_data |>
  group_by(h_geocode) |> # Group by home census block
  summarise(goods_p = sum(SI01), # Goods
            trade_p = sum(SI02), # Trade
            serve_p = sum(SI03), # Service
            total_p = sum(S000)) |> # Total 
  rename(geocode = h_geocode)

# Calculate the total number of trip attractions
total_attr <- oh_data |>
  group_by(w_geocode) |> # Group by home census block 
  summarize(goods_a = sum(SI01), # Goods
            trade_a = sum(SI02), # Trade
            serve_a = sum(SI03), # Service
            total_a = sum(S000)) |> # Total
  rename(geocode = w_geocode)

# Calculate trip generations (trip productions + trip attractions)
trip_gen <- full_join(total_prod,
                      total_attr) |>
  replace_na(list(goods_p = 0, 
                  goods_a = 0,
                  trade_p = 0,
                  trade_a = 0,
                  serve_p = 0,
                  serve_a = 0,
                  total_p = 0,
                  total_a = 0))

head(trip_gen)
```

# Load spatial data

```{r, warning = FALSE, message = FALSE}
# Download the census block boundaries for the counties of interest

# Define county FIPS (3-digit codes)
oh_counties_3_digit <- c("029") 

# Load block shapefiles 
msa_blocks <- blocks(state = "OH",
                    county = oh_counties_3_digit,
                    progress_bar = FALSE)

# Plotting the map
ggplot(msa_blocks) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(fill = NA, color = "orange") +
  theme_void()
```

# Creating `trip_gen_locs`

We will right join the trip generation table to the census blocks of interest (only include the ones in the trip generation table)

```{r}
trip_gen_locs <- msa_blocks |>
  rename(geocode = GEOID20) |>
  right_join(trip_gen) |>
  select(geocode, 
         goods_p, 
         trade_p, 
         serve_p,
         total_p,
         goods_a, 
         trade_a,
         serve_a,
         total_a) |>
  st_transform("WGS84")

leaflet(trip_gen_locs) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "orange",
              fillColor = "orange",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = trip_gen_locs$geocode)

nrow(trip_gen_locs) # This number of blocks is indicative of the computational expense and how long it will take to skim the network 
```

# Load the network

We will use BBBike (a cycle route planner) to extract/download the OpenStreetMap network

Details

* Area: 'Salem, Columbiana County, Ohio, 44460, United States' covers 80 square km
* Format: Protocolbuffer (PBF)

I downloaded the network .pbf file from BBBike, saved it in a folder called `network` $\rightarrow$ in the code below, we will use r5r to save the .pbf file as two shape files in an empty folder I created called `data`

Note: I had to manually delete the cache from my laptop in order to regenerate the shape files after redoing the BBBike step 

It seems like the goal is to extract the smallest street network that overlaps with the trip generation zones because if the street network area is too large, I start getting some errors when trying to save the .pbf file as shape files

```{r, eval = FALSE}
# Setting to eval = FALSE so that we don't repeat this process again
salem_core <- setup_r5(
  data_path = here("P4_trip_distribution", "network"),
  overwrite = TRUE,
  verbose = TRUE
)

street_vis <- street_network_to_sf(salem_core)

street_lines <- street_vis$edges
street_pts <- street_vis$vertices

# Write/save as two shape files
st_write(street_lines,
         here("P4_trip_distribution",
              "data",
              "street-lines.shp"))

st_write(street_pts,
         here("P4_trip_distribution",
              "data",
              "street-pts.shp"))

stop_r5()
```

```{r, message = FALSE, warning = FALSE}
# Reading the two network shape files now that they're created
street_lines <- here("P4_trip_distribution",
                     "data",
                     "street-lines.shp") |>
  st_read()

street_pts <- here("P4_trip_distribution",
                     "data",
                     "street-pts.shp") |>
  st_read()
```

# Overlaying the street network and trip generation zones

We want to make sure that the street network (red) covers at least the full trip generation zones (green)

If this is not the case, we need to shift the target area and re-download the BBBike OpenStreetMap network (I had to do this)

```{r}
base_map <- get_tiles(street_lines,
                      provider = "CartoDB.Positron",
                      zoom = 8,
                      crop = TRUE)

ggplot() +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(data = trip_gen_locs,
          color = "palegreen3",
          fill = "palegreen") +
  geom_sf(data = street_lines,
          color =  "salmon") +
  theme_void()
```



